Interrogating the linear variable importance of local explanations of non-linear models with animated linear projections

### 2021-12-22 _paper.rmd
edits from Jieyang and grammerly.

---
title: "Interrogating the linear variable importance of local explanations of black-box models with animated linear projections"
shorttitle: Cheem
author:
#- name: "Nicholas Spyrison"
  # address: |
  #   | Monash University
  #   | Faculty of Information Technology
  #   | ORCiD: 0000-0002-8417-0212
  # email: \email{nicholas.spyrison@monash.edu}
  # orcid: 0000-0002-8417-0212
#- name: "Dianne Cook"
  # affiliation: Monash University
  # address: |
  #   | ORCiD: 000-0002-3813-7155
  # orcid: 000-0002-3813-7155
#- name: "Kimbal Marriott"
  # affiliation: Monash University
  # address: |
  #   | ORCiD: 0000-0002-9813-0377
  # orcid: 0000-0002-9813-0377
# keywords:
#   formatted: [XAI, explanable artificial intelligence, black-box models, local # explanations, grand tour, manual tour, DR, dimension reduction, R]
#   plain:     [XAI, explanable artificial intelligence, black-box models, local # explanations, grand tour, manual tour, DR, dimension reduction, R]
abstract: |
 Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of eXplainable AI (XAI). XAI attempts to shed light on these models, one such approach is the use of local explanations. A local explanation of a model gives a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation, and approximate data and this attribution space side-by-side with linked brushing. After identifying an observation of interest its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with a technique called the tour. This tour animates many projections over small changes in the projection basis. Doing so allows a user to visually explore the data space through the lens of this local explanation and interrogate its variable importance. The implementation of our framework is available as an __R__ package __cheem__ available on CRAN.
# documentclass: jdssv
classoption: article
output:
  bookdown::pdf_document2:
  #pdf_document:
    keep_tex: true
    toc: false
    #template: template/jdssv_template.tex
    number_sections: yes
    fig_caption: yes
bibliography: cheem_paper.bib
#appendix:
#   - "appendix-1.Rmd"
# title:
#   formatted: "Methods for understanding the variable importance of local explanations of black-box models"
#   plain:     "Methods for understanding the variable importance of local explanations of black-box models"
#   short:     "Use of the manual tour to visually interrogate local explanations"
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
 - \usepackage{setspace} ## for title page spacing
 - \usepackage{hyperref} ## for all sorts of linking
---
```{r include=FALSE, cache=FALSE}
require("knitr")
require("kableExtra")
require("magrittr")
## Work packages
require("cheem")
require("spinifex")
## chunk options
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = FALSE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = TRUE,
  cache.lazy = FALSE)
```

@@

<!-- Introduce explanatory vs predictive modeling -->
Breiman and Shmueli [@breiman_statistical_2001; @shmueli_explain_2010] introduce the idea of distinguishing modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose such as hypothesis testing, while _predictive_ modeling predicts new, out-of-sample, observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use has important implications for model selection and development.

<!-- XAI & interpretability crisis -->
Black-box models are becoming increasingly common, but not without their share of controversy and issues [@oneil_weapons_2016; @kodiyan_overview_2019]. Black-box models have been known to reflect common biases, including sex [@dastin_amazon_2018; @duffy_apple_2019], race [@larson_how_2016], and age [@diaz_addressing_2018]. Such issues occur when biases existent in the training data, the model picks up on this influence on the response variable, which is then built into the model. Another issue is data drift when new data is outside the support of latent or exogenous explanatory variables. Data drift can lead to worse predictions [@lazer_parable_2014; @salzberg_why_2014]. Such cases highlight the need to make models fair, accountable, ethical, and transparent, which has led to the movement of XAI [@adadi_peeking_2018; @arrieta_explainable_2020].

@@

<!-- SHAP -->
To illustrate our work, we apply the model-agnostic explanation SHAP [@strumbelj_efficient_2010; @strumbelj_explaining_2014]. The exact details of SHAP are tangent to the ideas of this work, but suffice it to say that SHAP approximates variable importance by taking the median importance over permutations of the explanatory variables. To be exact, we apply a variant that enjoys a lower computational complexity, known as tree SHAP [@lundberg_consistent_2018].
<!-- 2nd half is more application details; move lower? -->

<!-- ## Data visualization tours -->
In multivariate data visualization, a _tour_ [@asimov_grand_1985; @buja_grand_1986; @lee_state_2021] is a sequence of linear projections of data onto a lower-dimensional space, typically 1-3D. Tours are viewed as an animation over small changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the structural shape of the shadow change gleans insight into the shape and features of the object.

<!-- manual tours --> 
There are various types of tours, which are distinguished by the generation of projection bases. In a _manual_ tour [@cook_manual_1997; @spyrison_spinifex_2020], the path is defined by changing the contribution of a selected variable. <!--tours and models --> Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries [@wickham_visualizing_2015], and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path [@lee_pptree_2013; @da_silva_projection_2021].

<!-- Purposed approach -->
The proposed approach is to use the radial, manual tour to interrogate a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package __cheem__ with an interactive application to facilitate analysis. We give case studies of toy and modern datasets for both classification and regression tasks.

<!-- What-if & counterfactual analysis -->
The change in the projection basis might feel similar to counterfactual, what-if analysis, such as _ceteris paribus_ [@biecek_ceterisparibus_2020]. Latin for “other things held constant” or “all else unchanged” is a counterfactual analysis showing how an observation's prediction would change from a change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual, observed case by that varies contributions of the basis, essentially the orientation of the data object. Another difference is that the basis must maintain orthonormality. That is to say, when the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction.

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following section, [Local explanation statistics](#sec:explanations), covers the background of the local explanation, SHAP, and the traditional visuals produced from it. [Tours and the radial tour](#sec:tour) digs deeper into these animations of continuous linear projections. The section [Application Design](#sec:applicationdesign) discusses the layout of the application, how it facilitates analysis, preprocessing, and package infrastructure. The section [Case Studies](#sec:casestudies) illustrates several applications of this method. We conclude with a [Discussion](#sec:discussion) of the insights we draw from classification and regression tasks.

@@

<!-- reminder of local explanation -->
Consider a highly non-linear model. At face value, it is hard to say which variable(s) are sensitive to the crossing of a classification boundary or identifies which variables caused an observation to have a relatively extreme residual. Local explanations shed light on these cases by approximating linear variable importance in the vicinity of one observation.

<!-- Taxonomy of local explanations -->
Figure 6 of @arrieta_explainable_2020 gives comprehensive summarization of the taxonomy and literature of explanation techniques. This includes a large number of model-specific explanations such as deepLIFT, [@shrikumar_not_2016; @shrikumar_learning_2017] a popular recursive method for estimating importance in neural networks. There are a fewer number of model-agnostic explanations, of which LIME, [@ribeiro_why_2016] SHAP, [@lundberg_unified_2017] and their variants are popular.

<!-- Uses of local explanations -->
These instance-level explanations are used in a variety of ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating which pixels were important [@simonyan_deep_2014]. For instance, snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain variable additive contributions from model intercept to the observation's prediction [@ribeiro_why_2016].

@@

<!-- fifa example -->
To illustrate SHAP and its original use, explaining the difference between the intercept and an observation's prediction, we use soccer data from FIFA 2020 season [@leone_fifa_2020]. We have 5000 observations of nine skill measures (after aggregating highly correlated variables). A random forest model is fit to regress the log wages, in 2020 Euros, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). We expect to see a difference in the attribution of the variable importance across the two positions of the players.

<!-- Illustration -->
Figure \@ref(fig:shapdistrbd) shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more informative to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot [@gosiewska_ibreakdown_2019], where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry in the variable ordering, so we opt to fix the order to panel b), namely, by decreasing the sum of the SHAP values.

```{r shapdistrbd, echo=F, fig.cap = "Illustration of the distribution of SHAP attributions and a breakdown plot. From FIFA 2020 data, a random forest model regresses wages from nine skill attributes for a star offensive and defensive player. The players have very different wages, but b) shows the distributions of the attributions permuting over 25 permutations in the explanatory variables. The medians of these distributions are the final SHAP values. The variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances make sense in light of the player's position. c) Breakdown plots of the observations the explanation used to additively explain the difference between the model intercept and the observation prediction."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

In summary, this highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, we showed how two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In the following section, we will be using normalized explanations as the starting projection basis to interrogate the explanation further.

@@

<!-- tours intro -->
A data visualization _tour_ animates many linear projections over small changes in the basis. One of the key features of the tour is the object permanence of the data points; one can track the relative changes of observations as the basis moves, as opposed to discretely jumping to an orthogonal view with no intermediate information. There  are various types of tours, which are distinguished by the selection of their basis paths [@lee_state_2021; @cook_grand_2008].

## Manual tours and the radial subset

<!-- Manual tour -->
The manual tour [@cook_manual_1997] defines its basis path by manipulating the basis contribution of a selected variable. A manipulation dimension is appended onto the projection plane, with a full contribution given to the selected variable. The bases are then selected based on rotating this newly created manipulation space. A crucial feature of the manual tour is that it allows users to control the variable contributions of the basis. This means that such manipulations can be selected and queued in advance or selected on the spot for human-in-the-loop analysis [@karwowski_international_2006]. However, this navigation is relatively time-consuming due to the huge volume of $p$-space. It is advisable to use this method to explore the sensitivity of the variable contribution to a previously identified feature of interest. In this case, the projection of the normalized explanations is the feature of interest.

<!-- radial tour -->
More generally, the manual tour can change the contribution of the variable in any of the display dimensions. We will with a specific more directed interaction, namely, a _radial_ tour. In a radial tour, the selected variable is allowed to change its magnitude of contribution but not its angle; it must move along the direction of its original contribution radius.

# Application design {#sec:applicationdesign}

Below we illustrate the two primary displays of the application: the global view and the tour view. Then we cover what we take away from the classification and regression tasks. Lastly, we discuss the preprocessing that before display.

@@

<!-- Purpose -->
The global view provides an essential context of all observations and allows  exploration of the separability of the data- and attribution-spaces. While ultimately, its purpose is to facilitate the selection of a primary point and comparison point.

<!-- Approximations of the spaces, PC1:2 -->
An approximation of these spaces is given as the first two principal components of their respective spaces. This is shown side-by-side next to model information, the prediction, and the observed response variable. The orientation and magnitude  of the variables are inscribed on a unit circle. While a single 2D projection will rarely encompass all of the structure of higher-dimensional spaces, it is a reasonable summarization, given the real task at hand, the selection of observations to explore further.

<!-- Predicted by observed -->
It is insightful to explore these two approximations against a visual of the model; prediction by observation is also displayed. Linked brushing and preserved aesthetic features such as circling misclassified observations help link information from the different spaces together.

@@

<!-- a function of the obs selected above -->
The global view facilitated the selection of a primary and optional comparison observation. The variable-level attribution of primary observation is normalized and used as the initial 1D basis in a radial tour. This is an approximation of  the contributions of the linear variables that best explain the difference between the model intercept and an observations prediction, not the local shape of the model surface.

@@

<!-- tour animation -->
The radial tour creates a basis path by varying the contribution of a selected variable, fully into and out of a projection frame. Doing so tests an individual variable's sensitivity to the structure identified by the local explanation. The default variable selected has the largest discrepancy between the primary and comparison observations attribution. In the following sections, we elaborate on the takeaways we draw from applying this approach in classification and regression tasks, respectively.

<!-- Segue to classification -->
Now that we have introduced the global view and corresponding cheem radial tour, let's discuss the differences between the classification and regression cases.

@@

What information do we glean from using this method on a classification task? Typically we select a misclassified observation compared to a correctly classified point that is nearby in data space. We start by seeing the data projected through the linear attribution, the combination that best justifies that prediction. By default, the manual tour varies the contribution of the variable with the largest difference between the primary and comparison observation. That is, we can test the sensitivity of each variable to structure identified by the local explanation; we are exploring the support of the explanation, evaluating the support or robustness of the prediction.

```{r classificationcase, echo=F, out.width = "100%", fig.cap = "Display illustrating the classification case. Both views are colored on predicted class and while red circles identify misclassified observations. The radial tour is a 1D projection starting at the normalized tree SHAP values of the primary point. The first frame is the linear-variable importances that best describe the difference from model intercept to this observation's prediction. We probe the support of the variable contributions by selecting a variable to vary the contribution."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

The regression case, we opt to color the global view on a statistic, the observation's residual, the log Mahanabis distance in data space (a measure of outlyingness), and the correlation of the attribution projection with the observed response. In the radial tour, the horizontal positions are the same, the  attribution projection, then varying as the radial tour changes basis. While the vertical position is fixed to the observed response variable and residuals for the middle and right panels, respectively. This changes the display from 1D density to a scatterplot. The basis is still one component, the horizontal position, independent of the vertical position.

```{r regressioncase, echo=F, out.width = "100%", fig.cap = "Display of the regression case. The global view can be colored on the correlation of the attribution projection and observed response. In the tour, the horizontal values are the same as the classification case. The vertical position is fixed to the observed response and residuals in the middle and right facets, respectively."}
knitr::include_graphics("./figures/app_regression.PNG")
```

@@

<!-- reactive vs exploration interactions -->
The application has several reactive selections that affect the data, coloring, and inputs for the tour to improve flexibility by extending the use cases. It also has interactive features in the global view that aid the analysis.

<!-- Interaction except -->
A tooltip displays while the cursor hovers over a point displays the observation number/name and classification information if appropriate. Linked brushing allows for the selection of points (with left click and drag) where those points will be highlighted in both plots. The information corresponding to the selected points is populated on a dynamic table. These interactions aid exploration of the spaces and ultimately the selection of the primary and comparison observations.

```{r interactions, echo=F, out.width = "100%", fig.cap = "Illustration of data explorations interactions in the global view. This view has linked brushing of the points where observations selected in one facet are highlighted in the other facets and populate an interactive tabular display below. Tooltips display when hovering over a point"}
knitr::include_graphics("./figures/app_interactions.PNG")
```

@@

The benefit of having dynamic interaction with data is predicated on a reasonably short render time. It is important to preprocess expensive operations so the application resources can be used efficiently. The work remaining at runtime is solely responding to inputs and rendering of visuals and tables. Below we discuss the steps and details of the reprocessing.

(ref:citeRf) @liaw_classification_2002
(ref:citeTs) @kominsarczyk_treeshap_2021
\begin{itemize}
	\item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
	\item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [(ref:citeRf)] for compatibility with the local explanation, which requires tree-based models.
	\item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, that is applicable to tree-based models. This is done with the package \textbf{treeshap} [(ref:citeTs), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

>>>>

---
title: "Interrogating the linear variable importance of local explanations of non-linear models with animated linear projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
#- name: "Nicholas Spyrison"
#     | Monash University
#     | Faculty of Information Technology
#   address: |
#     | ORCiD: 0000-0002-8417-0212
#   email: \email{nicholas.spyrison@monash.edu}
#   orcid: 0000-0002-8417-0212
#- name: "Dianne Cook"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 000-0002-3813-7155
#   orcid: 000-0002-3813-7155
#- name: "Kimbal Marriott"
#   affiliation: Monash University
#   address: |
#     | ORCiD: 0000-0002-9813-0377
#   orcid: 0000-0002-9813-0377
abstract: |
 Artificial Intelligence (AI) has seen a revitalization in recent years from the use of increasingly hard-to-interpret black-box models. In such models, increased predictive power comes at the cost of opaque factor analysis, which has led to the field of eXplainable AI (XAI). XAI attempts to shed light on these models. One such approach is _local explanations_. A local explanation of a model gives a point-estimate of linear variable importance in the vicinity of one observation. We extract explanations for each observation and approximate data, and this attribution space side-by-side with linked brushing. After identifying an observation of interest, its local explanation is used as a 1D projection basis. We then manipulate the magnitude of the variable contributions with the _tour_ technique. The tour animates many projections over small changes in the projection basis as it changes the contribution of one variable. Doing so allows a user to visually explore the data space through the lens of this local explanation and test its support. The implementation of our framework is available as an __R__ package __cheem__ available on CRAN.
# documentclass: jdssv
classoption: article
output:
  bookdown::pdf_document2:
  #pdf_document:
    keep_tex: true
    toc: false
    #template: template/jdssv_template.tex
    number_sections: yes
    fig_caption: yes
bibliography: cheem_paper.bib
#appendix:
# title:
#   formatted: "Methods for understanding the variable importance of local explanations of black-box models"
#   plain:     "Methods for understanding the variable importance of local explanations of black-box models"
#   short:     "Use of the manual tour to visually interrogate local explanations"
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
 - \usepackage{setspace} ## for title page spacing
 - \usepackage{hyperref} ## for all sorts of linking
---
```{r include=FALSE, cache=FALSE}
require("knitr")
require("kableExtra")
require("magrittr")
## Work packages
require("cheem")
require("spinifex")
## chunk options
knitr::opts_chunk$set(
  fig.align  = "center",
  echo       = FALSE,
  collapse   = TRUE,
  message    = FALSE,
  warning    = FALSE,
  error      = FALSE,
  cache      = TRUE,
  cache.lazy = FALSE)
```

@@

<!-- Introduce explanatory vs predictive modeling -->
Breiman and Shmueli [@breiman_statistical_2001; @shmueli_explain_2010] introduce the idea of distinguishing modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose such as hypothesis testing, while _predictive_ modeling predicts new, out-of-sample, observations. This distinction draws attention to the divide between interpretable models and black-box models. The intended use has important implications for model selection and development. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for more accurate non-linear models, the additional hindrance to interpretation should not be taken lightly.

<!-- XAI & interpretability crisis -->
Interpretability aside, black-box models are becoming increasingly common, but not without their share of controversy [@oneil_weapons_2016; @kodiyan_overview_2019]. Black-box models have been known to reflect common biases, including sex [@dastin_amazon_2018; @duffy_apple_2019], race [@larson_how_2016], and age [@diaz_addressing_2018]. Such issues occur when biases existent in the training data, the model picks up on this influence on the response variable, which is then built into the model. Another issue is data drift when new data is outside the support of latent or exogenous explanatory variables. Data drift can lead to worse predictions [@lazer_parable_2014; @salzberg_why_2014]. While typically more resistant to data drift, linear models will similarly find biases in the data. However, their transparency goes a long way into exploring and mitigating the biases. Such cases highlight the need to make models fair, accountable, ethical, and transparent, which has led to the movement of XAI [@adadi_peeking_2018; @arrieta_explainable_2020].

@@

<!-- Data visualization tours -->
In multivariate data visualization, a _tour_ [@asimov_grand_1985; @buja_grand_1986; @lee_state_2021] is a sequence of linear projections of data onto a lower-dimensional space, typically 1-3D. Tours are viewed as an animation over minor changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the shape of the shadow change conveys information of the structure and features of the object.

<!-- Manual tours --> 
There are various types of tours distinguished by the generation of projection bases. In a _manual_ tour [@cook_manual_1997; @spyrison_spinifex_2020], the path is defined by changing the contribution of a selected variable. <!--tours and models --> Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries [@wickham_visualizing_2015], and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path [@lee_pptree_2013; @da_silva_projection_2021].

<!-- Purposed approach -->
The proposed approach uses the radial, manual tour to interrogate a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package __cheem__ with an interactive application to facilitate analysis. We provide toy and modern datasets case studies for both classification and regression tasks.

<!-- What-if & counterfactual analysis -->
The change in the projection basis might feel similar to counterfactual, what-if analysis, such as _ceteris paribus_ [@biecek_ceterisparibus_2020]. This phrase, Latin for “other things held constant” or “all else unchanged”, shows how an observation's prediction would change from a marginal change in one explanatory variable given that other variables are held constant. It ignores correlations of the variables and imagines a case that was not observed. In contrast, our approach is a geometric explanation of the factual; it varies contributions of the variables by rotating the basis, a reorientation of the data object. Another difference is that the basis must remain orthonormal. That is to say, when the contribution of one variable decreases, the contributions of others necessarily increase such that there is a complete component in that direction.

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following section, [Local explanation statistics](#sec:explanations), covers the background of the local explanation, SHAP, and the traditional visuals produced from it. [Tours and the radial tour](#sec:tour) digs deeper into these animations of continuous linear projections. The section [Application Design](#sec:applicationdesign) discusses the visual layouts, how they facilitates analysis, data preprocessing, and package infrastructure. The section [Case Studies](#sec:casestudies) illustrates several applications of this method. We conclude with a [Discussion](#sec:cheemdiscussion) of the insights we draw from classification and regression tasks.

@@

<!-- Reminder of local explanation -->
Consider a highly non-linear model. At face value, it is hard to say which variable(s) are sensitive to  crossing a classification boundary or identify which variables caused an observation to have a relatively extreme residual. Local explanations shed light on these cases by approximating linear variable importance in the vicinity of one observation.

<!-- Taxonomy of local explanations -->
Figure six of @arrieta_explainable_2020 gives comprehensive summarization of the taxonomy and literature of explanation techniques. The figure includes a large number of model-specific explanations such as deepLIFT, [@shrikumar_not_2016; @shrikumar_learning_2017] a popular recursive method for estimating importance in neural networks. There are fewer model-agnostic explanations, of which LIME, [@ribeiro_why_2016] SHAP, [@lundberg_unified_2017] and their variants are popular.

<!-- Uses of local explanations -->
These instance-level explanations are used in various ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating which pixels were necessary [@simonyan_deep_2014]. For instance, snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain variable additive contributions from the model intercept to the observation's prediction [@ribeiro_why_2016].

@@

<!-- Fifa example -->
To illustrate SHAP and its original use, explaining the difference between the intercept and an observation's prediction, we use soccer data from FIFA 2020 season [@leone_fifa_2020]. We have 5000 observations of nine skill measures (after aggregating highly correlated variables). A random forest model is fit to regress the log wages, in 2020 Euros, from the skill measures. We then extract the SHAP values of a star offensive player (L. Messi) and defensive player (V. van Dijk). We expect to see a difference in the attribution of the variable importance across the two positions of the players.

```{r shapdistrbd, echo=F, fig.align='center', fig.cap = "Illustration of the distribution of SHAP attributions and a break-down plot. From FIFA 2020 data, a random forest model regresses wages from nine skill attributes for a star offensive and defensive player. The players have very different salaries, but a) shows the distributions of 25 permutations in the explanatory variables. The medians of these distributions are the final SHAP values. The variable importance differs across the exogenous information of player position. These explanations make sense; the variable importances seem realistic given the player's positions. b) Break-down plots of the observations using their explanations to additively cover the difference between the model intercept and the observation predictions."}
knitr::include_graphics("./figures/shap_distr_bd.png")
```

<!-- Illustration -->
Figure \@ref(fig:shapdistrbd) shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offense and movement skills are more important for the offensive player, while defensive and power skills are more informative to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel b) shows a simplified break-down plot [@gosiewska_ibreakdown_2019], where a local explanation is used to additively explain the difference from the intercept to the observation's prediction. Such additive approaches will show asymmetry in the variable ordering, so we opt to fix the order to panel a), by decreasing sum of the SHAP values.

<!-- Segue -->
In summary, this figure highlights how local explanations bring interpretability to a model, at least in the vicinity of their observations. In this instance, two players with different positions receive different profiles of variable importance to explain the prediction of their wages. In application we apply _tree SHAP_, a variant of SHAP that enjoys a lower computational complexity [@lundberg_consistent_2018]. Tree SHAP only compatible with tree-based models; we illustrate on random forests. In the following section, we will use normalized explanations as the starting projection basis to further interrogate the explanation.

@@

<!-- Tours intro -->
A data visualization _tour_ animates many linear projections over small changes in the basis. One of the critical features of the tour is the object permanence of the data points; one can track the relative changes of observations as the basis moves, as opposed to discretely jumping to an orthogonal view with no intermediate information. There are various types of tours that are distinguished by selecting their basis paths [@lee_state_2021; @cook_grand_2008].

## Manual tours and its radial case

<!-- Manual tour -->
The manual tour [@cook_manual_1997] defines its basis path by manipulating a selected variable's basis contribution. A manipulation dimension is appended onto the projection plane, giving a full contribution to the chosen variable. The bases are then selected based on rotating this newly created manipulation space. A crucial feature of the manual tour is that it allows users to control the variable contributions of the basis. Such manipulations can be selected and queued in advance or selected on the spot for human-in-the-loop analysis [@karwowski_international_2006]. However, this navigation is relatively time-consuming due to the vast volume of the display-space. It is advisable to use this method to explore the sensitivity of the variable contribution to a previously identified feature of interest. In this case, the projection of the normalized explanations is the feature of interest.

<!-- Radial tour -->
More generally, the manual tour can change the contribution of a variable to the display dimensions. We will apply a more directed interaction, namely, a _radial_ tour. In a radial tour, the selected variable is allowed to change its magnitude of contribution but not its angle; it must move along the direction of its original contribution.

# cheem viewer {#sec:applicationdesign}

Below we illustrate the two primary displays of the cheem viewer application: the global view and the tour view. Then we cover what we take away from the classification and regression tasks. Lastly, we discuss the preprocessing the data before application runtime.

@@

<!-- Purpose -->
The global view provides an essential context of all observations and facilitates the exploration of the separability of the data- and attribution-spaces. While comparison of these spaces is interesting, the global view's purpose is to facilitate the selection of observations. The local explanation of these points will be explored in more detail.

<!-- Approximations of the spaces, PC1:2 -->
An approximation of these spaces is given as the first two principal components of their respective spaces. Model information, observed response by its prediction, is also provided. The orientation and magnitude of the variables are inscribed on a unit circle. <!-- Interactions -->Misclassified observations are circled in red if applicable. Linked brushing between the plots and tabular display of select points facilitates exploration of the spaces and the model. A single 2D projection will not encompass all of the structure of higher-dimensional space. However, it is a reasonable summarization given the real task at hand; the selection of observations to explore further.

@@

<!-- A function of the obs selected -->
The global view facilitated the selection of a primary and optional comparison observation. The variable-level attribution of the primary observation is normalized and used as the initial 1D basis in a radial tour. This is an approximation of the contributions of the linear variables that best explain the difference between the model intercept and an observation's prediction, not the local shape of the model surface.

@@

<!-- Tour animation -->
The radial tour creates a basis path by varying the contribution of a selected variable, fully into and out of a projection frame. Doing so tests an individual variable's sensitivity to the structure identified by the local explanation.The default variable selected has the largest discrepancy between the attribution of primary and comparison observations. The following sections elaborate on the takeaways we draw from applying this approach in classification and regression tasks, respectively. <!-- Segue to classification -->Now that we have introduced the global view and corresponding cheem radial tour let us discuss the differences between the classification and regression cases.

@@

What information do we glean from using this method on a classification task? Typically we select a misclassified observation compared to a correctly classified point nearby in data space. The initial frame is the linear attribution of that observation's local explanation. By default, the manual tour varies the contribution of the variable with the largest difference between the primary and comparison observation. That is, we can test the sensitivity of each variable to structure identified by the local explanation; we are exploring the support of the explanation, evaluating the support or robustness of the prediction.

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Display illustrating the classification case. Plots are  colored on predicted class and red circles indicate misclassified observations. The radial tour is a 1D projection starting at the normalized tree SHAP values of the primary point. The first frame is the linear-variable importances that best describe the difference from model intercept to this observation's prediction. We probe the support of the variable contributions by selecting a variable to vary the contribution."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

In the regression case, the global view can be colored on a statistic to highlight the structure in the explanation space, including residuals, log Mahalanobis distance of data space (a measure of outlyingness), and the correlation of the attribution projection with the observed response. In the radial tour, the horizontal positions are the same, the basis projection of the radial tour. The vertical position is fixed to the observed response variable and residuals in the middle and right panels, respectively. Correspondingly, the display changes from univariate density to 2D scatterplot. The basis is still one component (horizontal) independent of the vertical position.

```{r regressioncase, echo=F, out.width = "100%", fig.cap="Display of the regression task. The global view can be colored on the correlation of the attribution projection and observed response. In the tour, the horizontal values are the same as the classification case; the projection through the basis. The vertical position is now maped to the observed y and residuals."}
knitr::include_graphics("./figures/app_regression.PNG")
```

@@

<!-- Reactive vs exploration interactions -->
The application has several reactive inputs that affect the data used, aesthetic display, and tour manipulation. These reactive inputs make the software flexible and extensible. The application also has It also has more exploratory interactions to help link points in the data and extract other information not on plot axes.

<!-- Exploratory interactions -->
A tooltip displays observation number/name and classification information while the cursor hovers over a point. Linked brushing allows the selection of points (left click and drag) where those points will be highlighted across plots. The information corresponding to the selected points is populated on a dynamic table. These interactions aid exploration of the spaces and, finally, identification of a primary and comparison observation.

```{r interactions, echo=F, out.width = "100%", fig.cap = "Illustration of data explorations interactions in the global view. This view has linked brushing of the points where observations selected in one facet are highlighted in the other facets and populate an interactive tabular display below. Tooltips display when hovering over an obersvation."}
knitr::include_graphics("./figures/app_interactions.PNG")
```

@@

It is vital to mitigate the render time of visuals, especially when users may want to iterate many times. In order to keep render time low all of the computational operations should be prepared before runtime. The work remaining at runtime is solely reacting to inputs and rendering of visuals and tables. Below we discuss the steps and details of the reprocessing.

(ref:citeRf) @liaw_classification_2002
(ref:citeTs) @kominsarczyk_treeshap_2021
\begin{itemize}
	\item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
	\item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [(ref:citeRf)] for compatibility with the local explanation, which requires tree-based models.
	\item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, compatible with tree-based models. Tree SHAP was calculated with the package \textbf{treeshap} [(ref:citeTs), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

<!-- Note on time of execution -->
The time to preprocess the data will vary significantly with the model and local explanation. For reference, the FIFA data, 5000 observations of nine explanatory variables, took 2.9 seconds to fit a random forest model of modest hyperparameters. Extracting the tree SHAP values of each observation took 254 seconds combined. PCA and statistics of the variables and attributions took 0.6 seconds. These runtimes were from a non-parallelized R session on a modern laptop, but suffice to say that the bulk of the time will be spent on the local attribution. An increased of model complexity or data dimensionality will quickly become an obstacle. This makes tree SHAP, with its reduced computational complexity, a good candidate to start with. Alternatively, the package __fastshap__ [@greenwell_fastshap_2020] claims extremely low runtimes, which are attributed to fewer calls to the prediction function, partial implementation in C++, and efficient use of logical subsetting.
